\texttt{ensmallen}: a flexible C++ library for efficient function optimization

### 2018-10-15 doc/mlsys_paper/paper.tex

@@

Optimization is often computationally intensive and may correspond
to most of the time taken to train a machine learning model.  For instance, the
training of deep neural networks is dominated by the optimization of the model
parameters on the data~\cite{schmidhuber2015deep}.
Even popular machine learning models such as logistic regression
have training times mostly dominated by an optimization~\cite{kingma2015adam}.
% TODO: might be nice to have something kind of anecdotal like 'even new
% students to the field of machine learning quickly encounter optimization' and
% cite, e.g., Andrew Ng's coursera course or some ML textbook or similar
%% CS: i think we don't need to explore this too much; better to cut out all the
%% CS: fat and stick with concrete examples, instead of veering off on tangents
%
% or maybe just a note about how many optimization techniques get published at
% NIPS every year?
%% CS: NIPS is too self-referential here

@@

\vspace*{-0.3em}
\section{Templates for Speed}
%\section{Automatic Metaprogramming for Efficiency Gains}
\vspace*{-0.5em}

%% CS: this sections is very weak and not convincing; it basically states
%% CS: "look, we can do inlines", but provides no real evidence
%% CS: and/or citations where the supposed efficiency gains are demonstrated.
%% CS: i suggest removing this section entirely,
%% CS: as it causes more harm than good (from a reviewer's point of view).
%% 
%% CS: in practice, effect of C++ function inlining is only notably
%% CS: pronounced when dealing with data comprised of lots of small vectors.
%% 
%% CS: for this section to be fleshed out, it would require reasoning based
%% CS: on citations covering compiler optimisation, effect of data size,
%% CS: experiments showing the effect of various vector sizes
%% CS: (eg. with and without inlining).
%% CS: there is no room in this paper to do this properly,
%% CS: and doing a poor job will reduce likelihood of acceptance.

Many optimization toolkits depend on the user providing a pointer to the
function that needs to be optimized.  For instance, the {\tt scipy.optimize}
package in Python can be used like this:

@@

During the call to {\tt minimize()}, the {\tt rosen()} function will be called
repeatedly---but it will be called as a function pointer, which must be
dereferenced each time {\tt rosen()} is called.  In addition, some
optimizations such as inlining (and many optimizations enabled by inlining) are
not possible when using function pointers.  Since many optimization algorithms
need to compute the objective numerous times, the overhead of the lookup and
missed optimization opportunities may be non-negligible.  This applies even for
SGD-like optimizers that train machine learning models on very large datasets.

%% CS: this is uninteresting and basically looks like filler
Since {\tt ensmallen} is written in C++ with templates, the function pointer
dereferencing overhead can be avoided and the compiler can choose to inline the
function call and perform additional optimizations when possible, such as
temporary variable elimination and dead code pruning.

>>>>

% fix the default super-ugly coloured link boxes (citations, urls, figure numbers, footnote numbers, etc)
\usepackage[pagebackref=false,breaklinks=true,colorlinks,bookmarks=false,urlcolor=blue,linkcolor=blue,citecolor=blue]{hyperref}

@@

Optimization is often computationally intensive and may correspond
to most of the time taken to train a machine learning model.  For instance, the
training of deep neural networks is dominated by the optimization of the model
parameters on the data~\cite{schmidhuber2015deep}.
Even popular machine learning models such as logistic regression
have training times mostly dominated by an optimization procedure~\cite{kingma2015adam}.
% TODO: might be nice to have something kind of anecdotal like 'even new
% students to the field of machine learning quickly encounter optimization' and
% cite, e.g., Andrew Ng's coursera course or some ML textbook or similar
%% CS: i think we don't need to explore this too much; better to cut out all the
%% CS: fat and stick with concrete examples, instead of veering off on tangents
%
% or maybe just a note about how many optimization techniques get published at
% NIPS every year?
%% CS: NIPS is too self-referential here

@@

\vspace*{-0.3em}
\section{Templates for speed}
\vspace*{-0.5em}

Many optimization toolkits depend on the user passing in a function pointer to a
function that needs to be optimized.  For instance, the {\tt scipy.optimize}
package in Python can be used like this:

@@

During the call to {\tt minimize()}, the {\tt rosen()} function will be called
repeatedly---but it will be called as a function pointer, which must be
dereferenced each time {\tt rosen()} is called.  In addition, some
optimizations such as inlining (and many optimizations enabled by inlining) are
not possible when using function pointers.  Since some optimization algorithms
need to compute the objective very many times, the overhead of the lookup and
missed optimization opportunities may be non-negligible.  This applies even for
SGD-like optimizers that train machine learning models on very large datasets!

Since {\tt ensmallen} is written in C++ with templates, the function pointer
dereferencing overhead can be avoided and the compiler can choose to inline the
function call and perform additional optimizations when possible, such as
temporary variable elimination and dead code pruning.
